---
title: "data-exploration"
format: html
---
## Data Exploration 
This data comes from the Smithsonian Environmental Research Center's Coastal Carbon Network project. More information on data files here: https://smithsonian.github.io/CCN-Community-Resources/soil_carbon_guidance.html 

### Environment loading
```{r}
library(tidyverse)
library(here)
library(dplyr)
library(ggplot2)
```

### Data Import
```{r}
# Import 'CCN_sites', 'CCN_species', 'CCN_cores', 'CCN_depth_series'

sites <- read_csv(here("CCN_data/CCN_sites.csv"))
species <- read_csv(here("CCN_data/CCN_species.csv"))
cores <- read_csv(here("CCN_data/CCN_cores.csv"))
depth_series <- read_csv(here("CCN_data/CCN_depthseries.csv"))

```

### Exploring the data set
```{r}
# Explore 'sites' data
glimpse(sites) # columns of interest: site_id (chr), vegetation_class, salinity_class
unique(sites) # what is emergent grass?
unique(sites$site_id) # some sites with odd IDs 
unique(sites$vegetation_class)

# Explore 'species' data
glimpse(species) # columns of interest: site_id, species_code, habitat
unique(species$site_id)
unique(species$habitat) 

# Explore 'cores' data
glimpse(cores) # columns of interest: vegetation class, habitat, 
unique(cores$vegetation_class)

# Explore 'depth_series'
glimpse(depth_series) # site_id is encoded, columns of interest: fraction_organic_matter, fraction_carbon, c14_age_se

```

### Data wrangling
1. I want to first filter the data to only the columns I want from each dataset. Keeping the column I plan to join the data by. I may not use the 'cores' data as it contains repetions of the 'vegetation_class' and 'habitat' columns. 

- sites: 'site_id', 'study_id', 'vegetation_class'
- species: 'site_id', 'study_id', 'species_code', 'habitat'
- depth_series: 'site_id', 'study_id', 'fraction_organic_matter', 'fraction_carbon', 'dry_bulk_density', 'c14_age', 
- cores: 'elevation', 'max_depth', 'core_id'

My predictor variables include 'habitat' (categorical), 'vegetation_class' (categorical), 'max depth' (continuous) and my response variables include 'fraction_carbon', 'fraction_organic_matter', 'species_code', 'dry_bulk_density', and 'c14_age'. 

#### Filter individual datasets 
```{r}
# sites
sites <- sites %>% 
  select( 'site_id', 'study_id', 'vegetation_class') %>% 
  drop_na()

# species

species <- species %>% 
  select('site_id', 'study_id', 'species_code', 'habitat') %>% 
  drop_na()

# depth_series (LOOK INTO: dropping na's removes all obs)
depth_series <- depth_series %>% 
  select('site_id', 'study_id', 'fraction_organic_matter', 'fraction_carbon', 'dry_bulk_density', 'c14_age')


# cores
cores <- cores %>% 
  select('site_id', 'study_id','elevation', 'max_depth', 'core_id' ) %>% 
  drop_na()

```

### Join data 
Joined data is quite large, brainstorming ways to further filter the data. 
```{r}
#joined_df <- full_join(sites, species, depth_series, by = c('site_id', 'study_id')) -- Can't join three at once 

joined_df <- sites %>%
  full_join(species, by = c('site_id', 'study_id')) %>%
  full_join(depth_series, by = c('site_id', 'study_id')) %>% 
  full_join(cores, by =  c('site_id', 'study_id')) 

glimpse(joined_df)
```
## Preliminary Data Visualizations
```{r, fig.cap= "A breakdown of habitat types within the joined dataset. This is helpful to know that marsh is a well-documented habitat type."}

joined_df %>% 
  ggplot(aes(habitat)) +
  geom_bar(fill = "#345084FF") +
  labs(x = "Habitat", 
       title = "Habitat Types") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

```

```{r}
#joined_df %>% 
  #ggplot(aes(x = habitat, y = fraction_carbon)) +
  #geom_boxplot(fill = "habitat") 
  #theme_bw()
```

```{r}
joined_df %>% 
  ggplot(aes(x = dry_bulk_density, y = fraction_carbon)) +
  geom_point(color = "#345084FF") +
  xlim(0, 1.5) + 
  ylim(0, 1) +
  labs(x = "Soil dry bulk density (g/cm^3)", 
       y = "Fraction carbon (dimensionless)", 
       title = "Relationship Between Soil Bulk Density and Fraction Carbon") +
  theme_bw()
```

Relationship between soil bulk density and fraction carbon: This is a relationship I was interested in, however from the view of the scatter plot, the data needs to be aggregated or filtered further to piece out a relationship. 

## Plans for the data
- turn habitat into a continuous variable
- dig deeper into question you want to ask with the dataset 
